---
title: "SomaScan v4.1 Proteomics: Arterial vs. Venous and Covariate Analysis"
author: "Gregory A. Chinn"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    fig-width: 11
    fig-height: 7
editor: visual
---

## Overview

SomaScan v4.1 PharmaServices run (PLT6574, 2022-02-26) on the same 24 EDTA plasma samples as the Luminex 65-plex study: 12 patients undergoing elective surgery, one simultaneous arterial (`-4`) and venous (`-2`) draw per patient. 7,596 aptamers measured; 7,481 pass ColCheck QC.

**Analyses:**

1.  Sample QC and PCA
2.  Arterial vs. venous: paired limma, 7,481 proteins
3.  Gene-set enrichment (GSEA) on the A-V ranking
4.  Covariate analysis on venous samples (obesity, hypothermia, hyperglycemia, sex)
5.  A-V gradient by covariate
6.  Cross-check against Luminex cytokine hits

------------------------------------------------------------------------

## Setup

```{r}
#| label: setup
#| message: false
#| warning: false

library(SomaDataIO)
library(limma)
library(tidyverse)
library(readxl)
library(ggplot2)
library(ggrepel)
library(pheatmap)
library(RColorBrewer)
library(fgsea)
library(msigdbr)

# Resolve project directory
data_dir <- tryCatch(
  dirname(rstudioapi::getActiveDocumentContext()$path),
  error = function(e) "/Users/gregchinn/Desktop/coding projects/Re_ Metabolomics"
)

# Colour palette used throughout
col_av    <- c("Arterial" = "#c0392b", "Venous" = "#2980b9")
col_yn    <- c("Yes" = "#c0392b", "No"  = "#2980b9")
col_surg  <- setNames(
  RColorBrewer::brewer.pal(3, "Set2"),
  c("Head/Neck", "Laparoscopic", "Spine")
)
```

## 1. Load Data and QC

```{r}
#| label: load-adat

# Use the anmlSMP file — hybNorm → medNorm → plateScale → calibration → ANML-QC → ANML-sample
adat_file <- file.path(
  data_dir,
  "SS-229807_v4.1_EDTA_Plasma.hybNorm.medNormInt.plateScale.calibration.anmlQC.qcCheck.anmlSMP.adat"
)
soma <- read_adat(adat_file)

cat(sprintf("Total rows: %d | Total aptamers: %d\n", nrow(soma), length(getAnalytes(soma))))
cat("Sample types:\n"); print(table(soma$SampleType))
```

```{r}
#| label: qc-flags

# Analyte-level QC
analyte_info <- getAnalyteInfo(soma)
cat("Aptamer ColCheck:\n"); print(table(analyte_info$ColCheck))

# Sample-level QC
samples_all <- soma |> dplyr::filter(SampleType == "Sample")
cat("\nSample RowCheck:\n"); print(table(samples_all$RowCheck))

flagged_samples <- samples_all |>
  dplyr::filter(RowCheck == "FLAG") |>
  select(SampleId, RowCheck)

cat("\nFlagged sample IDs:\n")
print(flagged_samples)
```

```{r}
#| label: build-analysis-set

# Keep only PASS analytes for all downstream analyses
pass_analytes <- analyte_info |>
  dplyr::filter(ColCheck == "PASS") |>
  pull(AptName)

cat("Analytes retained (ColCheck PASS):", length(pass_analytes), "\n")

# Keep all 24 patient samples; flag the 4 that failed RowCheck
soma_samples <- samples_all |>
  mutate(
    patient = str_extract(SampleId, "\\d{4}"),
    draw    = if_else(str_ends(SampleId, "-2"), "Venous", "Arterial"),
    flagged = RowCheck == "FLAG"
  ) |>
  arrange(patient, draw)

# log2-transform RFU for all downstream analyses
expr_all <- soma_samples |>
  select(all_of(pass_analytes)) |>
  as.matrix() |>
  log2()

rownames(expr_all) <- soma_samples$SampleId

cat(sprintf("Expression matrix: %d samples × %d analytes\n",
            nrow(expr_all), ncol(expr_all)))
cat("Flagged samples (included but noted):",
    sum(soma_samples$flagged), "\n")
```

### Load covariates

```{r}
#| label: load-covariates

surgery_map <- tribble(
  ~patient, ~surgery_type,              ~surgery_group,
  "0179",   "Spine",                    "Spine",
  "0193",   "Lap. cholecystectomy",     "Laparoscopic",
  "0209",   "Spine",                    "Spine",
  "0215",   "Spine",                    "Spine",
  "0223",   "Craniectomy",              "Head/Neck",
  "0245",   "Lap. adrenalectomy",       "Laparoscopic",
  "0250",   "Lap. nephrectomy",         "Laparoscopic",
  "0254",   "Carotid endarterectomy",   "Head/Neck",
  "0268",   "Thyroidectomy",            "Head/Neck",
  "0275",   "Transsphenoidal",          "Head/Neck",
  "0284",   "Spine",                    "Spine",
  "0291",   "Spine",                    "Spine"
)

clean_yn <- function(x) {
  case_when(
    str_to_upper(str_trim(x)) == "Y" ~ TRUE,
    str_to_upper(str_trim(x)) == "N" ~ FALSE,
    TRUE ~ NA
  )
}

covariate_raw <- read_xlsx(
  file.path(data_dir,
    "SS-229807_v4.1_EDTA_Plasma - Art vs Venous and surgery type and other covariates after AP2 PK copy.xlsx")
) |>
  rename(
    sample_id     = SampleId,
    art_iv_same   = `Art&IV on same side (Y/N)`,
    white_race    = `White Race (Y/N)`,
    male          = `Male (Y/N)`,
    obese         = `Obese (BMI>30)`,
    hypothermia   = `Hypothermia (Temp <36C)`,
    hyperglycemia = `Hyperglycemia (BG>150)`
  ) |>
  mutate(
    patient = str_extract(sample_id, "\\d{4}"),
    across(c(art_iv_same, white_race, male, obese, hypothermia, hyperglycemia), clean_yn)
  ) |>
  group_by(patient) |>
  summarise(across(c(art_iv_same, white_race, male, obese, hypothermia, hyperglycemia),
                   first),
            .groups = "drop") |>
  left_join(surgery_map, by = "patient")

soma_samples <- soma_samples |>
  left_join(covariate_raw, by = "patient")
```

------------------------------------------------------------------------

## 2. PCA — Sample-Level Overview

```{r}
#| label: pca
#| fig-height: 6

pca_res <- prcomp(expr_all, scale. = TRUE)
pct_var <- round(100 * pca_res$sdev^2 / sum(pca_res$sdev^2), 1)

pca_df <- as_tibble(pca_res$x[, 1:4]) |>
  mutate(
    SampleId      = soma_samples$SampleId,
    patient       = soma_samples$patient,
    draw          = soma_samples$draw,
    surgery_group = soma_samples$surgery_group,
    flagged       = soma_samples$flagged
  )

p1 <- ggplot(pca_df, aes(PC1, PC2, color = draw, shape = surgery_group,
                          label = patient)) +
  geom_line(aes(group = patient), color = "gray70", linewidth = 0.5) +
  geom_point(aes(size = flagged), alpha = 0.9) +
  geom_text_repel(size = 2.5, show.legend = FALSE) +
  scale_color_manual(values = col_av) +
  scale_shape_manual(values = c("Head/Neck" = 17, "Laparoscopic" = 15, "Spine" = 16)) +
  scale_size_manual(values = c("FALSE" = 3, "TRUE" = 5), guide = "none") +
  labs(
    title    = "PCA: SomaScan 7,481 proteins (log2 RFU, scaled)",
    subtitle = "Lines connect arterial–venous pairs per patient. Large points = RowCheck FLAG.",
    x        = paste0("PC1 (", pct_var[1], "%)"),
    y        = paste0("PC2 (", pct_var[2], "%)"),
    color    = "Draw", shape = "Surgery"
  ) +
  theme_bw(base_size = 12) +
  theme(legend.position = "right")

p2 <- ggplot(pca_df, aes(PC3, PC4, color = draw, shape = surgery_group,
                          label = patient)) +
  geom_line(aes(group = patient), color = "gray70", linewidth = 0.5) +
  geom_point(size = 3, alpha = 0.9) +
  geom_text_repel(size = 2.5, show.legend = FALSE) +
  scale_color_manual(values = col_av) +
  scale_shape_manual(values = c("Head/Neck" = 17, "Laparoscopic" = 15, "Spine" = 16)) +
  labs(
    x = paste0("PC3 (", pct_var[3], "%)"),
    y = paste0("PC4 (", pct_var[4], "%)"),
    color = "Draw", shape = "Surgery"
  ) +
  theme_bw(base_size = 12) +
  theme(legend.position = "right")

print(p1)
print(p2)
```

```{r}
#| label: pca-loadings

# Top drivers of PC1 and PC2
loadings <- as_tibble(pca_res$rotation[, 1:2], rownames = "AptName") |>
  left_join(analyte_info |> select(AptName, Target, EntrezGeneSymbol), by = "AptName")

cat("Top PC1 drivers (positive):\n")
loadings |> slice_max(PC1, n = 10) |>
  select(Target, EntrezGeneSymbol, PC1) |> print()

cat("\nTop PC1 drivers (negative):\n")
loadings |> slice_min(PC1, n = 10) |>
  select(Target, EntrezGeneSymbol, PC1) |> print()
```

------------------------------------------------------------------------

## 3. Arterial vs. Venous — Paired limma

**Design:** block on patient (12 levels), coefficient of interest = draw (Arterial vs. Venous). Equivalent to a paired t-test across 7,481 proteins simultaneously. BH-FDR correction.

```{r}
#| label: limma-av

patient_f <- factor(soma_samples$patient)
draw_f    <- factor(soma_samples$draw, levels = c("Venous", "Arterial"))

design_av <- model.matrix(~ patient_f + draw_f)

fit_av  <- lmFit(t(expr_all), design_av)
fit_av  <- eBayes(fit_av, trend = TRUE, robust = TRUE)

av_results <- topTable(fit_av, coef = "draw_fArterial",
                        number = Inf, sort.by = "P") |>
  rownames_to_column("AptName") |>
  as_tibble() |>
  left_join(
    analyte_info |> select(AptName, Target, TargetFullName,
                            EntrezGeneSymbol, EntrezGeneID, UniProt, Dilution),
    by = "AptName"
  ) |>
  mutate(
    direction = if_else(logFC > 0, "Arterial > Venous", "Venous > Arterial"),
    sig_FDR   = adj.P.Val < 0.05,
    sig_nom   = P.Value   < 0.05
  ) |>
  arrange(adj.P.Val)

cat(sprintf("Proteins tested: %d\n", nrow(av_results)))
cat(sprintf("Nominal (p < 0.05): %d\n",     sum(av_results$sig_nom)))
cat(sprintf("FDR-significant (BH < 0.05): %d\n", sum(av_results$sig_FDR)))
cat(sprintf("\nDirection (FDR hits): "))
if (sum(av_results$sig_FDR) > 0) {
  print(av_results |> dplyr::filter(sig_FDR) |> count(direction))
} else {
  cat("none at FDR < 0.05\n")
}

cat("\nTop 30 by adjusted p-value:\n")
av_results |>
  select(Target, EntrezGeneSymbol, logFC, AveExpr, t, P.Value, adj.P.Val, direction) |>
  mutate(across(c(logFC, AveExpr, t, P.Value, adj.P.Val), ~ signif(., 3))) |>
  print(n = 30)
```

### Volcano plot

```{r}
#| label: soma-volcano
#| fig-height: 8

# Label top 25 by p-value
label_df <- av_results |> slice_min(P.Value, n = 25)

ggplot(av_results, aes(x = logFC, y = -log10(P.Value),
                        color = direction, alpha = sig_nom)) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "gray50") +
  geom_vline(xintercept = 0, color = "gray80") +
  geom_point(size = 1.2) +
  geom_text_repel(data = label_df, aes(label = Target),
                  size = 2.5, max.overlaps = 30, box.padding = 0.35,
                  show.legend = FALSE) +
  scale_color_manual(values = c("Arterial > Venous" = "#c0392b",
                                 "Venous > Arterial" = "#2980b9")) +
  scale_alpha_manual(values = c("TRUE" = 1, "FALSE" = 0.25), guide = "none") +
  labs(
    title    = "SomaScan: Arterial vs. Venous (paired limma, N = 12, 7,481 proteins)",
    subtitle = "Dashed = nominal p < 0.05. Top 25 labeled.",
    x        = "log2FC (Arterial / Venous)",
    y        = "-log10(p-value)",
    color    = NULL
  ) +
  theme_bw(base_size = 12) +
  theme(legend.position = "bottom")
```

### MA plot

```{r}
#| label: soma-ma

ggplot(av_results, aes(x = AveExpr, y = logFC, color = sig_nom)) +
  geom_point(size = 0.8, alpha = 0.5) +
  geom_hline(yintercept = 0, color = "gray40") +
  geom_text_repel(data = av_results |> slice_min(P.Value, n = 20),
                  aes(label = Target), size = 2.5, max.overlaps = 20,
                  show.legend = FALSE) +
  scale_color_manual(values = c("FALSE" = "gray70", "TRUE" = "#c0392b"),
                     labels = c("p ≥ 0.05", "p < 0.05")) +
  labs(
    title = "MA plot: SomaScan A-V",
    x = "Average log2 expression (AveExpr)",
    y = "log2FC (Arterial / Venous)",
    color = NULL
  ) +
  theme_bw(base_size = 12) +
  theme(legend.position = "bottom")
```

### Top hit paired trajectories

```{r}
#| label: soma-paired-dots
#| fig-height: 8

top_apts <- av_results |> slice_min(P.Value, n = 12) |> pull(AptName)
top_tgts <- av_results |> slice_min(P.Value, n = 12) |>
  mutate(label = if_else(is.na(Target) | Target == "", AptName, Target)) |>
  pull(label)

names(top_tgts) <- top_apts

soma_samples |>
  select(patient, draw, surgery_group, all_of(top_apts)) |>
  mutate(across(all_of(top_apts), ~ log2(.))) |>
  pivot_longer(all_of(top_apts), names_to = "AptName", values_to = "log2_rfu") |>
  mutate(
    draw    = factor(draw, levels = c("Venous", "Arterial")),
    protein = top_tgts[AptName],
    protein = factor(protein, levels = top_tgts)
  ) |>
  ggplot(aes(x = draw, y = log2_rfu, group = patient, color = surgery_group)) +
  geom_line(alpha = 0.5) +
  geom_point(size = 2) +
  facet_wrap(~ protein, scales = "free_y", ncol = 4) +
  scale_color_manual(values = col_surg) +
  labs(
    title = "Top 12 proteins: paired arterial vs. venous per patient",
    x = NULL, y = "log2 RFU", color = "Surgery"
  ) +
  theme_bw(base_size = 10) +
  theme(legend.position = "bottom")
```

### Heatmap: top 50 proteins

```{r}
#| label: soma-heatmap
#| fig-height: 14
#| fig-width: 9

top50_apts <- av_results |> slice_min(P.Value, n = 50) |> pull(AptName)

hm_mat <- t(expr_all[, top50_apts])
colnames(hm_mat) <- soma_samples$SampleId

# Row labels = protein name
rownames(hm_mat) <- av_results |>
  dplyr::filter(AptName %in% top50_apts) |>
  arrange(match(AptName, top50_apts)) |>
  mutate(label = if_else(is.na(Target) | Target == "", AptName, Target)) |>
  pull(label)

# Column annotation
col_anno <- data.frame(
  Draw    = soma_samples$draw,
  Surgery = soma_samples$surgery_group,
  Flagged = ifelse(soma_samples$flagged, "Yes", "No"),
  row.names = soma_samples$SampleId
)
anno_colors <- list(
  Draw    = col_av,
  Surgery = col_surg,
  Flagged = c("Yes" = "orange", "No" = "white")
)

pheatmap(
  hm_mat,
  scale            = "row",
  annotation_col   = col_anno,
  annotation_colors = anno_colors,
  show_colnames    = FALSE,
  cluster_rows     = TRUE,
  cluster_cols     = TRUE,
  color            = colorRampPalette(rev(brewer.pal(9, "RdBu")))(100),
  fontsize_row     = 7,
  main             = "Top 50 proteins by A-V p-value (row-scaled log2 RFU)"
)
```

------------------------------------------------------------------------

## 4. Gene-Set Enrichment Analysis (GSEA)

Rank all 7,481 PASS proteins by their limma t-statistic (positive = higher in arterial). Run fgsea against MSigDB Hallmark gene sets (H) and KEGG pathways (C2:CP:KEGG). Gene IDs from SomaScan EntrezGeneID annotation.

```{r}
#| label: gsea-prep

# Build named ranking vector: EntrezGeneID → t-statistic
# Some aptamers share an Entrez ID (multiple aptamers per protein) → keep max |t|
rank_df <- av_results |>
  dplyr::filter(!is.na(EntrezGeneID), EntrezGeneID != "") |>
  mutate(EntrezGeneID = as.character(EntrezGeneID)) |>
  group_by(EntrezGeneID) |>
  slice_max(abs(t), n = 1, with_ties = FALSE) |>
  ungroup()

ranked_vec <- setNames(rank_df$t, rank_df$EntrezGeneID)
ranked_vec <- sort(ranked_vec, decreasing = TRUE)

cat("Unique Entrez gene IDs in ranking:", length(ranked_vec), "\n")
```

```{r}
#| label: gsea-hallmark
#| message: false

set.seed(42)

h_sets <- msigdbr(species = "Homo sapiens", category = "H") |>
  select(gs_name, entrez_gene) |>
  mutate(entrez_gene = as.character(entrez_gene)) |>
  group_by(gs_name) |>
  summarise(genes = list(entrez_gene), .groups = "drop")

hallmark_list <- setNames(h_sets$genes, h_sets$gs_name)

gsea_hallmark <- fgsea(
  pathways  = hallmark_list,
  stats     = ranked_vec,
  minSize   = 10,
  maxSize   = 500,
  nPermSimple = 10000
) |>
  as_tibble() |>
  arrange(pval)

cat("Hallmark GSEA — nominal hits (p < 0.05):", sum(gsea_hallmark$pval < 0.05), "\n\n")
gsea_hallmark |>
  dplyr::filter(pval < 0.05) |>
  select(pathway, NES, pval, padj, size) |>
  mutate(across(c(NES, pval, padj), ~ signif(., 3))) |>
  print(n = 30)
```

```{r}
#| label: gsea-kegg
#| message: false

kegg_sets <- msigdbr(species = "Homo sapiens", category = "C2", subcategory = "CP:KEGG_LEGACY") |>
  select(gs_name, entrez_gene) |>
  mutate(entrez_gene = as.character(entrez_gene)) |>
  group_by(gs_name) |>
  summarise(genes = list(entrez_gene), .groups = "drop")

kegg_list <- setNames(kegg_sets$genes, kegg_sets$gs_name)

gsea_kegg <- fgsea(
  pathways    = kegg_list,
  stats       = ranked_vec,
  minSize     = 10,
  maxSize     = 500,
  nPermSimple = 10000
) |>
  as_tibble() |>
  arrange(pval)

cat("KEGG GSEA — nominal hits (p < 0.05):", sum(gsea_kegg$pval < 0.05), "\n\n")
gsea_kegg |>
  dplyr::filter(pval < 0.05) |>
  select(pathway, NES, pval, padj, size) |>
  mutate(across(c(NES, pval, padj), ~ signif(., 3))) |>
  print(n = 30)
```

### GSEA enrichment plot: top pathways

```{r}
#| label: gsea-barplot
#| fig-height: 8

all_gsea <- bind_rows(
  gsea_hallmark |> mutate(collection = "Hallmark"),
  gsea_kegg     |> mutate(collection = "KEGG")
) |>
  dplyr::filter(pval < 0.05) |>
  mutate(
    short_name  = str_replace_all(pathway, "HALLMARK_|KEGG_", "") |>
                  str_replace_all("_", " ") |>
                  str_to_title(),
    direction   = if_else(NES > 0, "Enriched in Arterial", "Enriched in Venous")
  ) |>
  arrange(NES)

if (nrow(all_gsea) > 0) {
  ggplot(all_gsea, aes(x = NES, y = reorder(short_name, NES),
                        fill = direction, alpha = -log10(pval))) +
    geom_col() +
    geom_vline(xintercept = 0, color = "gray30") +
    scale_fill_manual(values = c("Enriched in Arterial" = "#c0392b",
                                  "Enriched in Venous"   = "#2980b9")) +
    scale_alpha_continuous(range = c(0.4, 1)) +
    facet_wrap(~ collection, scales = "free_y") +
    labs(
      title = "GSEA: A-V ranking — nominal hits (p < 0.05)",
      x = "Normalized Enrichment Score (positive = enriched in arterial)",
      y = NULL, fill = NULL
    ) +
    theme_bw(base_size = 11) +
    theme(legend.position = "bottom")
} else {
  cat("No pathways reach nominal significance.\n")
}
```

------------------------------------------------------------------------

## 5. Covariate Analysis — Protein Levels by Patient Characteristics

**Method:** Venous samples only (N = 12). For each binary covariate, fit a limma linear model (expression \~ covariate) and apply eBayes. Present top hits and a summary heatmap. All results are hypothesis-generating; N = 12 is underpowered for 7,481-analyte testing.

```{r}
#| label: soma-cov-setup

# Venous samples only
ven_idx      <- soma_samples$draw == "Venous"
soma_ven     <- soma_samples[ven_idx, ]
expr_ven     <- expr_all[ven_idx, ]

binary_covariates <- c("obese", "hypothermia", "hyperglycemia", "male")

run_soma_covariate <- function(expr_mat, sample_df, cov_name, analyte_info) {
  group  <- sample_df[[cov_name]]
  keep   <- !is.na(group)

  if (sum(keep) < 6 || length(unique(group[keep])) < 2) {
    message("Skipping ", cov_name, ": insufficient non-NA observations")
    return(NULL)
  }

  design_cov <- model.matrix(~ group[keep])
  fit_cov    <- lmFit(t(expr_mat[keep, ]), design_cov)
  fit_cov    <- eBayes(fit_cov, trend = TRUE, robust = TRUE)

  topTable(fit_cov, coef = 2, number = Inf, sort.by = "P") |>
    rownames_to_column("AptName") |>
    as_tibble() |>
    left_join(analyte_info |> select(AptName, Target, EntrezGeneSymbol, UniProt),
              by = "AptName") |>
    mutate(
      covariate = cov_name,
      n_pos     = sum(group[keep] == TRUE,  na.rm = TRUE),
      n_neg     = sum(group[keep] == FALSE, na.rm = TRUE)
    ) |>
    arrange(P.Value)
}

cov_results_list <- map(
  binary_covariates,
  ~ run_soma_covariate(expr_ven, soma_ven, .x, analyte_info)
)
names(cov_results_list) <- binary_covariates

cov_results_all <- bind_rows(cov_results_list)

cat("Nominal covariate hits (p < 0.05) per covariate:\n")
cov_results_all |>
  dplyr::filter(P.Value < 0.05) |>
  count(covariate) |>
  print()
```

### Top hits per covariate

```{r}
#| label: soma-cov-top

cov_results_all |>
  dplyr::filter(P.Value < 0.05) |>
  group_by(covariate) |>
  slice_min(P.Value, n = 10) |>
  select(covariate, Target, EntrezGeneSymbol, logFC, t, P.Value, adj.P.Val, n_pos, n_neg) |>
  mutate(across(c(logFC, t, P.Value, adj.P.Val), ~ signif(., 3))) |>
  ungroup() |>
  print(n = 80)
```

### Heatmap: –log10(p) across all covariates

```{r}
#| label: soma-cov-heatmap
#| fig-height: 14
#| fig-width: 9

hit_apts <- cov_results_all |>
  dplyr::filter(P.Value < 0.05) |>
  pull(AptName) |>
  unique()

if (length(hit_apts) > 0) {
  hm_cov <- cov_results_all |>
    dplyr::filter(AptName %in% hit_apts) |>
    mutate(
      label       = if_else(is.na(Target) | Target == "", AptName, Target),
      neg_log10_p = pmin(-log10(P.Value), 4)
    ) |>
    group_by(covariate, label) |>
    summarise(neg_log10_p = max(neg_log10_p), .groups = "drop") |>
    pivot_wider(names_from = covariate, values_from = neg_log10_p, values_fill = 0) |>
    column_to_rownames("label") |>
    as.matrix()

  pheatmap(
    hm_cov,
    color        = colorRampPalette(c("white", "#f7dc6f", "#c0392b"))(50),
    breaks       = seq(0, 4, length.out = 51),
    cluster_rows = TRUE,
    cluster_cols = FALSE,
    fontsize_row = 6,
    fontsize_col = 9,
    main         = "–log10(p) by protein and covariate (venous; capped at 4)\nAll exploratory: N = 12",
    border_color = NA,
    angle_col    = 45
  )
} else {
  cat("No nominal hits across covariates.\n")
}
```

### Surgery type: Kruskal-Wallis on top A-V hits

```{r}
#| label: soma-surgery
#| fig-height: 8

# Use top 9 A-V hits for surgery comparison (venous samples)
top9_apts  <- av_results |> slice_min(P.Value, n = 9) |> pull(AptName)
top9_tgts  <- av_results |> slice_min(P.Value, n = 9) |>
  mutate(label = if_else(is.na(Target) | Target == "", AptName, Target)) |>
  pull(label)
names(top9_tgts) <- top9_apts

kw_surgery <- map_dfr(top9_apts, function(apt) {
  y  <- expr_ven[, apt]
  kw <- kruskal.test(y ~ soma_ven$surgery_group)
  tibble(AptName = apt, p_kruskal = kw$p.value)
})

soma_ven |>
  select(patient, surgery_group, all_of(top9_apts)) |>
  pivot_longer(all_of(top9_apts), names_to = "AptName", values_to = "log2_rfu") |>
  left_join(kw_surgery, by = "AptName") |>
  mutate(protein = top9_tgts[AptName],
         protein = paste0(protein, "\n(KW p=", signif(p_kruskal, 2), ")"),
         protein = factor(protein)) |>
  ggplot(aes(x = surgery_group, y = log2_rfu, color = surgery_group)) +
  geom_jitter(width = 0.15, size = 2.5, alpha = 0.8) +
  stat_summary(fun = median, geom = "crossbar", width = 0.5,
               color = "black", linewidth = 0.5) +
  facet_wrap(~ protein, scales = "free_y", ncol = 3) +
  scale_color_manual(values = col_surg) +
  labs(
    title = "Top 9 A-V proteins by surgery type (venous, log2 RFU)",
    x = NULL, y = "log2 RFU", color = "Surgery"
  ) +
  theme_bw(base_size = 10) +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 30, hjust = 1))
```

------------------------------------------------------------------------

## 6. Arterial–Venous Gradient by Covariate

Compute the per-patient A-V delta (log2 arterial RFU − log2 venous RFU) for each protein. Test delta against each binary covariate (limma linear model; single-sample design — less stable than the A-V analysis; treat as exploratory).

```{r}
#| label: soma-delta

# Ensure arterial and venous are aligned by patient
art_soma <- soma_samples |> dplyr::filter(draw == "Arterial") |> arrange(patient)
ven_soma <- soma_samples |> dplyr::filter(draw == "Venous")   |> arrange(patient)

stopifnot(all(art_soma$patient == ven_soma$patient))

expr_art   <- expr_all[art_soma$SampleId, ]
expr_ven_m <- expr_all[ven_soma$SampleId, ]

delta_mat  <- expr_art - expr_ven_m       # log2 ratio (A - V) per protein
rownames(delta_mat) <- art_soma$patient

delta_meta <- art_soma |>
  select(patient, surgery_group, obese, hypothermia, hyperglycemia, male)

cat("A-V delta matrix:", nrow(delta_mat), "patients ×", ncol(delta_mat), "proteins\n")
```

```{r}
#| label: soma-delta-tests

run_delta_cov <- function(delta_expr, sample_df, cov_name, analyte_info) {
  group <- sample_df[[cov_name]]
  keep  <- !is.na(group)

  if (sum(keep) < 4 || length(unique(group[keep])) < 2) return(NULL)

  design_d <- model.matrix(~ group[keep])
  fit_d    <- lmFit(t(delta_expr[keep, ]), design_d)
  fit_d    <- eBayes(fit_d, trend = TRUE, robust = TRUE)

  topTable(fit_d, coef = 2, number = Inf, sort.by = "P") |>
    rownames_to_column("AptName") |>
    as_tibble() |>
    left_join(analyte_info |> select(AptName, Target, EntrezGeneSymbol),
              by = "AptName") |>
    mutate(covariate = cov_name) |>
    arrange(P.Value)
}

delta_cov_results <- map(
  binary_covariates,
  ~ run_delta_cov(delta_mat, delta_meta, .x, analyte_info)
) |>
  setNames(binary_covariates) |>
  bind_rows()

cat("Nominal A-V delta hits (p < 0.05) per covariate:\n")
delta_cov_results |>
  dplyr::filter(P.Value < 0.05) |>
  count(covariate) |>
  print()

cat("\nTop hits:\n")
delta_cov_results |>
  dplyr::filter(P.Value < 0.05) |>
  group_by(covariate) |>
  slice_min(P.Value, n = 5) |>
  select(covariate, Target, EntrezGeneSymbol, logFC, t, P.Value, adj.P.Val) |>
  mutate(across(c(logFC, t, P.Value, adj.P.Val), ~ signif(., 3))) |>
  ungroup() |>
  print(n = 60)
```

### Delta dot plots

```{r}
#| label: soma-delta-dots
#| fig-height: 10

top_delta_hits <- delta_cov_results |>
  dplyr::filter(P.Value < 0.10) |>
  group_by(covariate) |>
  slice_min(P.Value, n = 3) |>
  ungroup()

if (nrow(top_delta_hits) > 0) {
  delta_long <- as_tibble(delta_mat, rownames = "patient") |>
    pivot_longer(-patient, names_to = "AptName", values_to = "delta")

  delta_cov_long <- delta_meta |>
    pivot_longer(all_of(binary_covariates),
                 names_to = "covariate", values_to = "group_val")

  plot_delta <- top_delta_hits |>
    select(covariate, AptName) |>
    left_join(delta_long, by = "AptName") |>
    left_join(delta_cov_long, by = c("patient", "covariate")) |>
    left_join(analyte_info |> select(AptName, Target), by = "AptName") |>
    dplyr::filter(!is.na(group_val)) |>
    mutate(
      group_label = if_else(group_val, "Yes", "No"),
      protein     = if_else(is.na(Target) | Target == "", AptName, Target),
      panel_label = paste0(covariate, ": ", protein)
    )

  if (nrow(plot_delta) > 0) {
    print(
      ggplot(plot_delta, aes(x = group_label, y = delta, color = group_label)) +
        geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
        geom_jitter(width = 0.1, size = 2.5, alpha = 0.8) +
        stat_summary(fun = median, geom = "crossbar", width = 0.4,
                     color = "black", linewidth = 0.5) +
        facet_wrap(~ panel_label, scales = "free_y", ncol = 3) +
        scale_color_manual(values = col_yn) +
        labs(
          title = "A-V delta (log2 arterial − venous) by covariate — hits at p < 0.10",
          x = NULL, y = "log2 A-V delta", color = NULL
        ) +
        theme_bw(base_size = 10) +
        theme(legend.position = "none",
              strip.text = element_text(size = 7))
    )
  } else {
    cat("No plottable delta data after NA removal.\n")
  }
} else {
  cat("No A-V delta hits reach p < 0.10.\n")
}
```

------------------------------------------------------------------------

## 7. Full 65-Plex Luminex vs. SomaScan Comparison

### Re-run Luminex A-V analysis

```{r}
#| label: luminex-reload
#| message: false

# Plate map
plate_map <- tribble(
  ~unknown_id,  ~patient, ~draw,
  "Unknown1",   "0179",   "Venous",
  "Unknown2",   "0193",   "Venous",
  "Unknown3",   "0209",   "Venous",
  "Unknown4",   "0215",   "Venous",
  "Unknown5",   "0179",   "Arterial",
  "Unknown6",   "0193",   "Arterial",
  "Unknown7",   "0209",   "Arterial",
  "Unknown8",   "0215",   "Arterial",
  "Unknown9",   "0223",   "Venous",
  "Unknown10",  "0245",   "Venous",
  "Unknown11",  "0250",   "Venous",
  "Unknown12",  "0254",   "Venous",
  "Unknown13",  "0223",   "Arterial",
  "Unknown14",  "0245",   "Arterial",
  "Unknown15",  "0250",   "Arterial",
  "Unknown16",  "0254",   "Arterial",
  "Unknown17",  "0268",   "Venous",
  "Unknown18",  "0275",   "Venous",
  "Unknown19",  "0284",   "Venous",
  "Unknown20",  "0291",   "Venous",
  "Unknown21",  "0268",   "Arterial",
  "Unknown22",  "0275",   "Arterial",
  "Unknown23",  "0284",   "Arterial",
  "Unknown24",  "0291",   "Arterial"
)

lx_raw <- read_csv(
  file.path(data_dir, "AP_65plex_byJMR.csv"),
  skip = 54, show_col_types = FALSE
) |>
  dplyr::filter(str_starts(Sample, "Unknown"))

lx_cols <- lx_raw |> select(-Location, -Sample, -`Total Events`) |> names()

lx_avg <- lx_raw |>
  mutate(across(all_of(lx_cols), ~ suppressWarnings(as.numeric(.)))) |>
  left_join(plate_map, by = c("Sample" = "unknown_id")) |>
  group_by(patient, draw) |>
  summarise(across(all_of(lx_cols), ~ mean(., na.rm = TRUE)), .groups = "drop")

lx_art <- lx_avg |> dplyr::filter(draw == "Arterial") |> arrange(patient)
lx_ven <- lx_avg |> dplyr::filter(draw == "Venous")   |> arrange(patient)

lx_results <- map_dfr(lx_cols, function(cyt) {
  a <- lx_art[[cyt]]; v <- lx_ven[[cyt]]
  test <- tryCatch(wilcox.test(a, v, paired = TRUE, exact = FALSE),
                   error = function(e) list(p.value = NA))
  tibble(
    luminex_name  = cyt,
    log2FC_lx     = log2((mean(a, na.rm=TRUE) + 0.1) / (mean(v, na.rm=TRUE) + 0.1)),
    p_value_lx    = test$p.value
  )
}) |>
  mutate(p_adj_lx = p.adjust(p_value_lx, method = "BH"))

cat("Luminex results reloaded:", nrow(lx_results), "cytokines\n")
```

### Map all 65 panel names to gene symbols

```{r}
#| label: panel-map

# Full 65-plex → gene symbol mapping
luminex_gene_map <- tribble(
  ~luminex_name,          ~gene_symbol,
  "MCP-2(CCL8)",          "CCL8",
  "IL-2R",                "IL2RA",
  "MIP-1a(CCL3)",         "CCL3",
  "SDF-1a",               "CXCL12",
  "IL-27",                "IL27",
  "LIF",                  "LIF",
  "IL-1b",                "IL1B",
  "IL-2",                 "IL2",
  "IL-4",                 "IL4",
  "IL-5",                 "IL5",
  "IP-10(CXCL10)",        "CXCL10",
  "IL-6",                 "IL6",
  "IL-7",                 "IL7",
  "IL-8(CXCL8)",          "CXCL8",
  "IL-10",                "IL10",
  "BLC(CXCL13)",          "CXCL13",
  "EOTAXIN-2(CCL24)",     "CCL24",
  "EOTAXIN(CCL11)",       "CCL11",
  "IL-12p70",             "IL12A",
  "IL-13",                "IL13",
  "IL-17A(CTLA-8)",       "IL17A",
  "IL-31",                "IL31",
  "SCF",                  "KITLG",
  "G-CSF(CSF-3)",         "CSF3",
  "IFNg",                 "IFNG",
  "GM-CSF",               "CSF2",
  "TNFa",                 "TNF",
  "HGF",                  "HGF",
  "MIP-1b(CCL4)",         "CCL4",
  "IFNa",                 "IFNA1",
  "EOTAXIN-3(CCL26)",     "CCL26",
  "MCP-1(CCL2)",          "CCL2",
  "IL-9",                 "IL9",
  "MIF",                  "MIF",
  "TNFb",                 "LTA",
  "NGFb",                 "NGF",
  "MIP-3a(CCL20)",        "CCL20",
  "I-TAC(CXCL11)",        "CXCL11",
  "TRAIL",                "TNFSF10",
  "Fractalkine(CX3CL1)",  "CX3CL1",
  "GROa(CXCL1)",          "CXCL1",
  "IL-1a",                "IL1A",
  "IL-23",                "IL23A",
  "MMP-1",                "MMP1",
  "IL-15",                "IL15",
  "IL-18",                "IL18",
  "M-CSF",                "CSF1",
  "MCP-3(CCL7)",          "CCL7",
  "MIG(CXCL9)",           "CXCL9",
  "IL-16",                "IL16",
  "IL-21",                "IL21",
  "IL-3",                 "IL3",
  "CD40L",                "CD40LG",
  "FGF-2",                "FGF2",
  "IL-22",                "IL22",
  "VEGF-A",               "VEGFA",
  "TSLP",                 "TSLP",
  "IL-20",                "IL20",
  "ENA-78(CXCL5)",        "CXCL5",
  "CD30",                 "TNFRSF8",
  "TNF-RII",              "TNFRSF1B",
  "BAFF",                 "TNFSF13B",
  "MDC",                  "CCL22",
  "APRIL",                "TNFSF13",
  "TWEAK",                "TNFSF12"
)

cat("Panel map entries:", nrow(luminex_gene_map), "\n")
```

### Merge Luminex and SomaScan A-V results

```{r}
#| label: platform-merge

# For proteins with multiple aptamers, keep the one with the smallest p-value
soma_best <- av_results |>
  group_by(EntrezGeneSymbol) |>
  slice_min(P.Value, n = 1, with_ties = FALSE) |>
  ungroup()

platform_df <- luminex_gene_map |>
  left_join(lx_results, by = "luminex_name") |>
  left_join(
    soma_best |> select(EntrezGeneSymbol, logFC_soma = logFC,
                         t_soma = t, P.Value_soma = P.Value,
                         adj.P.Val_soma = adj.P.Val),
    by = c("gene_symbol" = "EntrezGeneSymbol")
  ) |>
  mutate(
    in_soma        = !is.na(logFC_soma),
    dir_lx         = if_else(log2FC_lx  > 0, "Art > Ven", "Ven > Art"),
    dir_soma       = if_else(logFC_soma > 0, "Art > Ven", "Ven > Art"),
    concordant     = in_soma & (dir_lx == dir_soma),
    sig_lx         = p_value_lx < 0.05,
    sig_soma       = P.Value_soma < 0.05
  )

cat("Luminex cytokines found in SomaScan:", sum(platform_df$in_soma), "of 65\n")
cat("Direction concordant:", sum(platform_df$concordant, na.rm=TRUE), "\n")
cat("Direction discordant:", sum(!platform_df$concordant & platform_df$in_soma, na.rm=TRUE), "\n\n")

platform_df |>
  dplyr::filter(in_soma) |>
  select(luminex_name, gene_symbol, log2FC_lx, logFC_soma,
         p_value_lx, P.Value_soma, concordant) |>
  mutate(across(c(log2FC_lx, logFC_soma, p_value_lx, P.Value_soma), ~ signif(., 3))) |>
  arrange(p_value_lx) |>
  print(n = 65)
```

### Scatter plot: Luminex log2FC vs SomaScan logFC

```{r}
#| label: platform-scatter
#| fig-height: 7

scatter_df <- platform_df |>
  dplyr::filter(in_soma) |>
  mutate(
    sig_category = case_when(
      sig_lx & sig_soma  ~ "Both p < 0.05",
      sig_lx             ~ "Luminex only",
      sig_soma           ~ "SomaScan only",
      TRUE               ~ "Neither"
    ),
    label_pt = if_else(sig_lx | sig_soma, gene_symbol, NA_character_)
  )

# Pearson r for concordance
r_val <- cor(scatter_df$log2FC_lx, scatter_df$logFC_soma,
             use = "complete.obs", method = "pearson")
r_sp  <- cor(scatter_df$log2FC_lx, scatter_df$logFC_soma,
             use = "complete.obs", method = "spearman")

ggplot(scatter_df, aes(x = log2FC_lx, y = logFC_soma,
                        color = sig_category, label = label_pt)) +
  geom_hline(yintercept = 0, color = "gray70") +
  geom_vline(xintercept = 0, color = "gray70") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray50") +
  geom_point(size = 3, alpha = 0.85) +
  geom_text_repel(size = 2.8, max.overlaps = 25, box.padding = 0.3,
                  na.rm = TRUE, show.legend = FALSE) +
  scale_color_manual(values = c(
    "Both p < 0.05"  = "#8e44ad",
    "Luminex only"   = "#e67e22",
    "SomaScan only"  = "#27ae60",
    "Neither"        = "gray60"
  )) +
  annotate("text", x = -Inf, y = Inf, hjust = -0.1, vjust = 1.3,
           label = sprintf("Pearson r = %.2f\nSpearman ρ = %.2f", r_val, r_sp),
           size = 3.5, color = "gray20") +
  labs(
    title    = "Platform concordance: Luminex vs. SomaScan A-V log2FC",
    subtitle = sprintf("%d of 65 cytokines matched in SomaScan. Dashed = identity line.",
                       sum(scatter_df$in_soma)),
    x        = "log2FC Luminex (Arterial / Venous, Wilcoxon)",
    y        = "log2FC SomaScan (Arterial / Venous, limma)",
    color    = NULL
  ) +
  theme_bw(base_size = 12) +
  theme(legend.position = "bottom")
```

### Direction concordance summary

```{r}
#| label: concordance-table

cat("Concordance table (cytokines matched in both platforms):\n")
platform_df |>
  dplyr::filter(in_soma) |>
  count(dir_lx, dir_soma) |>
  mutate(concordant = dir_lx == dir_soma) |>
  arrange(desc(concordant)) |>
  print()

cat("\nCytokines with discordant direction:\n")
platform_df |>
  dplyr::filter(in_soma, !concordant) |>
  select(luminex_name, gene_symbol, log2FC_lx, logFC_soma,
         p_value_lx, P.Value_soma) |>
  mutate(across(where(is.numeric), ~ signif(., 3))) |>
  print()
```

### Per-patient concordance: arterial vs. arterial, venous vs. venous

For each overlapping protein, compare the per-patient Luminex MFI to the SomaScan log2-RFU **within the same draw type**. Each scatter panel has 12 points (one per patient) per draw; Spearman r quantifies how well the two platforms rank patients relative to each other.

```{r}
#| label: per-patient-crosswalk
#| message: false

# Identify best SomaScan aptamer per gene (highest mean RFU among pass_analytes)
apt_mean_rfu <- soma_samples |>
  select(all_of(pass_analytes)) |>
  summarise(across(everything(), ~ mean(., na.rm = TRUE))) |>
  pivot_longer(everything(), names_to = "AptName", values_to = "mean_rfu")

soma_gene_best <- analyte_info |>
  select(AptName, Target, EntrezGeneSymbol) |>
  dplyr::filter(AptName %in% pass_analytes,
                EntrezGeneSymbol %in% luminex_gene_map$gene_symbol) |>
  left_join(apt_mean_rfu, by = "AptName") |>
  group_by(EntrezGeneSymbol) |>
  slice_max(mean_rfu, n = 1, with_ties = FALSE) |>
  ungroup()

# Final per-protein crosswalk: one row per matched analyte pair
per_pt_crosswalk <- luminex_gene_map |>
  dplyr::filter(luminex_name %in% lx_cols) |>
  left_join(soma_gene_best |> select(EntrezGeneSymbol, AptName, Target),
            by = c("gene_symbol" = "EntrezGeneSymbol")) |>
  dplyr::filter(!is.na(AptName))

cat("Matched analyte pairs for per-patient concordance:", nrow(per_pt_crosswalk), "\n")
```

```{r}
#| label: per-patient-build

# Luminex: MFI per patient × draw (already averaged duplicates in lx_avg)
lumi_long <- lx_avg |>
  select(patient, draw, all_of(per_pt_crosswalk$luminex_name)) |>
  pivot_longer(all_of(per_pt_crosswalk$luminex_name),
               names_to = "luminex_name", values_to = "mfi")

# SomaScan: log2-RFU per patient × draw
soma_long <- soma_samples |>
  select(patient, draw, all_of(per_pt_crosswalk$AptName)) |>
  mutate(across(all_of(per_pt_crosswalk$AptName), log2)) |>
  pivot_longer(all_of(per_pt_crosswalk$AptName),
               names_to = "AptName", values_to = "log2_rfu")

per_pt_df <- per_pt_crosswalk |>
  select(luminex_name, AptName, Target) |>
  left_join(lumi_long, by = "luminex_name") |>
  left_join(soma_long, by = c("AptName", "patient", "draw")) |>
  mutate(draw = factor(draw, levels = c("Venous", "Arterial")))
```

```{r}
#| label: per-patient-cor

per_pt_cor <- per_pt_df |>
  group_by(Target, luminex_name, draw) |>
  summarise(
    r_spearman = cor(mfi, log2_rfu, method = "spearman", use = "complete.obs"),
    n          = sum(!is.na(mfi) & !is.na(log2_rfu)),
    .groups    = "drop"
  )

cat("Per-patient Spearman r (Luminex MFI vs. SomaScan log2-RFU):\n")
per_pt_cor |>
  pivot_wider(
    id_cols     = c(Target, luminex_name),
    names_from  = draw,
    values_from = r_spearman,
    names_prefix = "r_"
  ) |>
  mutate(
    mean_r = rowMeans(across(starts_with("r_")), na.rm = TRUE),
    across(c(starts_with("r_"), mean_r), ~ round(., 3))
  ) |>
  arrange(desc(mean_r)) |>
  print(n = 70)
```

```{r}
#| label: per-patient-scatter
#| fig-height: 20
#| fig-width: 14

per_pt_df |>
  mutate(panel_label = paste0(Target, "\n(", luminex_name, ")")) |>
  ggplot(aes(x = mfi, y = log2_rfu, color = draw, shape = draw)) +
  geom_point(size = 2.5, alpha = 0.85) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 0.7) +
  facet_wrap(~ panel_label, scales = "free", ncol = 5) +
  scale_color_manual(values = col_av) +
  scale_shape_manual(values = c("Arterial" = 17, "Venous" = 16)) +
  labs(
    title    = "Per-patient concordance: Luminex MFI vs. SomaScan log2-RFU",
    subtitle = "Each point = one patient. Arterial vs. venous shown separately. Lines = per-draw linear fit.",
    x        = "Luminex MFI",
    y        = "SomaScan log2-RFU",
    color    = NULL, shape = NULL
  ) +
  theme_bw(base_size = 9) +
  theme(
    legend.position = "bottom",
    strip.text      = element_text(size = 7),
    axis.text       = element_text(size = 7)
  )
```

```{r}
#| label: per-patient-cor-bar
#| fig-height: 8
#| fig-width: 10

per_pt_cor |>
  mutate(Target = fct_reorder(Target, r_spearman, .fun = mean, .na_rm = TRUE)) |>
  ggplot(aes(x = Target, y = r_spearman, fill = draw)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6, alpha = 0.85) +
  geom_hline(yintercept = c(0, 0.7), linetype = c("solid", "dashed"),
             color = c("gray30", "gray60")) +
  scale_fill_manual(values = col_av) +
  scale_y_continuous(limits = c(-1, 1), breaks = seq(-1, 1, 0.25)) +
  coord_flip() +
  labs(
    title    = "Per-patient Spearman r: Luminex MFI vs. SomaScan log2-RFU",
    subtitle = "Dashed line = r = 0.70. Sorted by mean r across draws.",
    x        = NULL,
    y        = "Spearman r (N = 12 patients per draw)",
    fill     = NULL
  ) +
  theme_bw(base_size = 11) +
  theme(legend.position = "bottom")
```

------------------------------------------------------------------------

## 8. 65-Cytokine Gene Set Enrichment in SomaScan Ranking

Treat the 65 Luminex panel proteins as a custom gene set and test whether they are collectively enriched toward the arterial or venous pole of the SomaScan A-V t-statistic ranking. This tests whether the **cytokine/chemokine compartment as a whole** shows a directional A-V bias in the proteome-wide ranking.

```{r}
#| label: custom-gsea

# Build Entrez IDs for the 65-cytokine gene set from SomaScan annotations
# (avoids needing org.Hs.eg.db)
panel_entrez <- av_results |>
  dplyr::filter(EntrezGeneSymbol %in% luminex_gene_map$gene_symbol,
                !is.na(EntrezGeneID), EntrezGeneID != "") |>
  pull(EntrezGeneID) |>
  as.character() |>
  unique()

cat("Entrez IDs recovered for the 65-cytokine panel:", length(panel_entrez), "\n")

# Custom gene set list
custom_sets <- list(
  `Luminex_65plex_Cytokine_Panel` = panel_entrez
)

set.seed(42)
gsea_panel <- fgsea(
  pathways    = custom_sets,
  stats       = ranked_vec,
  minSize     = 5,
  maxSize     = 500,
  nPermSimple = 100000
) |>
  as_tibble()

cat("\nGSEA result for the 65-cytokine gene set:\n")
gsea_panel |>
  select(pathway, NES, pval, padj, size, leadingEdge) |>
  mutate(across(c(NES, pval, padj), ~ signif(., 3))) |>
  print()

cat("\nLeading edge proteins:\n")
le_symbols <- av_results |>
  dplyr::filter(EntrezGeneID %in% unlist(gsea_panel$leadingEdge)) |>
  pull(EntrezGeneSymbol) |>
  unique()
cat(paste(le_symbols, collapse = ", "), "\n")
```

### Enrichment plot

```{r}
#| label: gsea-panel-plot
#| fig-height: 5

plotEnrichment(custom_sets[["Luminex_65plex_Cytokine_Panel"]], ranked_vec) +
  labs(
    title    = "GSEA enrichment: Luminex 65-plex cytokine panel",
    subtitle = sprintf("NES = %.2f, p = %.3f — positive NES = collectively enriched in arterial",
                       gsea_panel$NES, gsea_panel$pval),
    x        = "Rank in SomaScan A-V t-statistic (left = more arterial)",
    y        = "Enrichment score"
  ) +
  theme_bw(base_size = 12)
```

### Where do the 65 cytokines sit in the SomaScan ranking?

```{r}
#| label: panel-rank-plot
#| fig-height: 6

panel_ranks <- av_results |>
  mutate(rank = row_number()) |>
  dplyr::filter(EntrezGeneSymbol %in% luminex_gene_map$gene_symbol) |>
  left_join(luminex_gene_map, by = c("EntrezGeneSymbol" = "gene_symbol")) |>
  left_join(lx_results |> select(luminex_name, log2FC_lx, p_value_lx),
            by = "luminex_name") |>
  mutate(
    sig_lx   = p_value_lx < 0.05,
    sig_soma = P.Value    < 0.05,
    label    = if_else(sig_lx | sig_soma, EntrezGeneSymbol, NA_character_)
  ) |>
  arrange(rank)

ggplot(panel_ranks, aes(x = rank, y = logFC,
                         color = sig_lx, shape = sig_soma,
                         label = label)) +
  geom_hline(yintercept = 0, color = "gray60") +
  geom_point(size = 3, alpha = 0.85) +
  geom_text_repel(size = 2.8, max.overlaps = 20, box.padding = 0.3,
                  na.rm = TRUE, show.legend = FALSE) +
  scale_color_manual(
    values = c("TRUE" = "#e67e22", "FALSE" = "gray60"),
    labels = c("TRUE" = "Luminex p < 0.05", "FALSE" = "Luminex p ≥ 0.05")
  ) +
  scale_shape_manual(
    values = c("TRUE" = 17, "FALSE" = 16),
    labels = c("TRUE" = "SomaScan p < 0.05", "FALSE" = "SomaScan p ≥ 0.05")
  ) +
  scale_x_continuous(
    sec.axis = sec_axis(~ . / nrow(av_results) * 100, name = "Percentile")
  ) +
  labs(
    title    = "Position of 65 Luminex cytokines in the SomaScan A-V ranking",
    subtitle = "Positive logFC = higher in arterial. Rank 1 = most arterial.",
    x        = "Rank in SomaScan proteome (1 = most arterial, 7481 = most venous)",
    y        = "SomaScan log2FC (Arterial / Venous)",
    color    = NULL, shape = NULL
  ) +
  theme_bw(base_size = 12) +
  theme(legend.position = "bottom")
```

------------------------------------------------------------------------

## Interpretation Notes

**Sample flags:** Four samples failed SomaScan RowCheck QC (`000_0193-2`, `000_0193-4`, `000_0245-4`, `000_0268-4`). These are included in the paired analysis because ANML normalization has already corrected for sample-level signal variation; excluding them would eliminate patient 0193 entirely and reduce paired power. Results for those specific patients should be interpreted with caution.

**Power:** N = 12 paired samples against 7,481 proteins. The paired design captures the within-patient A-V contrast efficiently, but FDR correction across this many proteins is very conservative at this sample size. Nominal p-value threshold (p \< 0.05) is the interpretively relevant cutoff; treat any FDR-significant finding as a robust signal and nominal hits as candidate proteins.

**GSEA interpretation:** GSEA on the limma t-statistic ranking is more powerful than individual protein testing because it aggregates signal across gene sets. Pathways enriched in arterial blood would suggest systemic arterial supply of those mediators; pathways enriched in venous return suggest peripheral tissue release.

**Cross-platform consistency:** IP-10/CXCL10 direction (venous \> arterial) is the strongest a priori hypothesis from the Luminex data. Consistent direction in SomaScan (despite different aptamer-based measurement) would strengthen that interpretation substantially.
